Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems over Hilbert Spaces

### 2023-05-14 latex/intro.tex
Some minor changes, plus an explanation of how to modify clustered to non clustered
Done?

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov2012, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in Section
\ref{section:example}). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

\begin{figure}[H]
    \centering
    \includegraphics[height=0.5\textwidth]{example.eps}
    \caption{Measurement clusterization for the inversion of the
      initial condition of the 1D heat equation. For each measurement,
      its index is plotted where the measurement is taken. Fmong fou
      locations are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis), for varying numbers of measurements
      (y-axis). Measurement locations (numbers) were chosen according
      to the D-optimality criterion. Measurement clusterization
      already occurs for four measurements: the third measurement is
      overlaid with the fourth. For five measurements, first and
      second measurements are clustered, as well as the fourth and the
      fifth.}
  \label{fig:clusterization illustration}
\end{figure}

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging ``close''
measurements \cite{fedorov2012}; however, this strategy merely
overlooks the phenomenon of measurement clusterization. An alternative
solution lies in \emph{clusterization-free design}s, where measurement
locations are deliberately chosen to be distant from one another. This
can be achieved by imposing distance constraints between measurements
or by introducing correlated errors that account for both observation
error and model misspecification \cite{Ucinski05}. For instance, in
the context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{hooker2009}.

In spatial problems involving choice of measuremnts within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements
from a coarse grid in $\Omega$ \cite{koval2020, alexanderian2021,
  attia2020, alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. This approach incurs a significant
computational cost as it requires solving a difficult combinatorial
optimization problem for measurements locations over a finite set. The
combinatorial optimization problem is usually relaxed by first
assigning optimal measurement ``weights'' in $\mathbb{R}_+$ to the
potential measurement locations. Some researchers incorporate a
sparsifying $\ell_1$ penalty term into the design criterion, which is
subsequently thresholded to achieve the desired binary design over the
coarse grid \cite{horesh2008borehole}. Others progressively relax the
$\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a
stochastic optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in $\Omega$ inherently results in a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, we recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, we strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves its objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does measurement clusterization occur?
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Should we aim to avoid measurement clusterization?
%
Is it possible to substitute an optimal clustered design with an
equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

@@

\begin{restatable}[D-optimal designs with vanishing model error]{theorem}{char}\label{thm:char}
  Let:
  \begin{itemize}
  \item $\hilp, \hilo$ Hilbert spaces.
  \item $\fwd:\hilp \to \hilo$ a linear compact operator.
  \item $\pr \sim \mathcal{N}(0, \prcov)$ prior Gaussian measure on $\hilp$,
    where $\prcov:\hilp \to \hilp$ is the prior covariance operator.
  \item $\obs: \hilo \to \mathbb{R}^m$ measurement operator, where $m
    \in \mathbb{N}$ is the number of measurements taken.
  \item $\sigma^2 \in \mathbb{R}_{+}$ observation noise variance,
    $\data = \obs \fwd \param + \eps$, where $\eps \in \mathbb{R}^m$
    is iid $\mathcal{N}(0, \sigma^2)$ noise.
  \item $\post$ the posterior measure, with covariance $\postcov$.
  \item A D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) :&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \sigma^{-2} \prcov^{1/2} \fwd ^*
        \obs^* \obs \fwd \prcov^{1/2}).
      \end{split}
    \end{align*}
  \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of $\fwd\prcov\fwd^*$
    in decreasing order of magnitude.
  %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
  \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.
  \item $\obs$ a D-optimal design operator
    \begin{equation*}
      \obs = \argmax_{\|\meas_j\| = 1, j=1,\dots,m}\tar(\obs).
    \end{equation*}
  \end{itemize}

@@

Other contributions include three useful lemmas: In Lemma
\ref{lemma:free} we decompose a symmetric positive definite matrix $M
\in \mathbb{R}^{k \times k}$ with $\ttr M = m \in \mathbb{N}$ as $M =
AA^t$, where $A$ has unit norm columns. In Lemma \ref{lemma:MDL} we
prove a matrix determinant lemma in Hilbert spaces. Finally, in Lemma
\ref{lemma:lax} we generalize a lemma for calculating
$\frac{\der}{\der t} \log \det (I + X(t))$, where $X(t)$ is an
operator valued function \cite{Lax07}.

>>>>

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov2012, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in the
supplementary material). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}[H]
    \centering
    \includegraphics[height=0.5\textwidth]{example.eps}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain $\Omega = [0, 1]$
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging close measurements
\cite{fedorov2012}; however, this strategy merely overlooks the
phenomenon of measurement clusterization. An alternative solution lies
in \emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{hooker2009}.

In spatial problems involving choice of measurements within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements
from a coarse grid in $\Omega$ \cite{koval2020, alexanderian2021,
  attia2020, alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. This approach incurs a significant
computational cost as it requires solving a difficult combinatorial
optimization problem for measurement locations over a finite set. The
combinatorial optimization problem is usually relaxed by first
assigning optimal measurement weights in $\mathbb{R}_+$ to the
potential measurement locations. Some researchers incorporate a
sparsifying $\ell_1$ penalty term into the design criterion, which is
subsequently thresholded to achieve the desired binary design over the
coarse grid \cite{horesh2008borehole}. Others progressively relax the
$\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a
stochastic optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in $\Omega$ fundamentally changes the optimal design
problem and thus results in a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, researchers recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, they strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves its objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does measurement clusterization occur?
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Should we aim to avoid measurement clusterization?
%
Is it possible to substitute an optimal clustered design with an
equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

@@

We first give some general definitions, and then state Theorem
\ref{thm:char}: Let $\hilp, \hilo$ Hilbert spaces, $\fwd:\hilp \to
\hilo$ a linear compact operator. Let $\obs: \hilo \to \mathbb{R}^m$ a
linear measurement operator, where $m \in \mathbb{N}$ is the number of
measurements taken. Let $\sigma^2 \in \mathbb{R}_{+}$ observation
noise variance, $\data = \obs \fwd \param + \eps$, where $\eps \in
\mathbb{R}^m$ is iid $\mathcal{N}(0, \sigma^2)$ noise. Let $\pr \sim
\mathcal{N}(\prmean, \prcov)$ prior Gaussian measure on $\hilp$, where
$\prcov$ is the prior covariance operator and let $\post$ the
posterior (Gaussian) measure.

\begin{restatable}{theorem}{char}\label{thm:char}
  Let:
  \begin{itemize}
    \item The D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) %:&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \sigma^{-2} \prcov^{1/2} \fwd ^*
        \obs^* \obs \fwd \prcov^{1/2}), 
      \end{split}
    \end{align*}
  \item $\obs$ a D-optimal design operator
    \begin{equation*}
      \obs = \argmax_{\|\meas_j\| = 1, j=1,\dots,m}\tar(\obs),
    \end{equation*}
  \item $\{\lambda_i\}_{i=1}^\infty$ eigenvalues of
    $\fwd\prcov\fwd^*$ in decreasing order of magnitude.
    %% \item $\{\ev_i\}_{i=1}^\infty$ their corresponding eigenvectors.
  \item $\{\eta_i\}_{i=1}^\infty$ eigenvalues of $\obs^*\obs$.

  \end{itemize}

@@

Lastly, in Lemma \ref{lemma:free} we decompose a symmetric positive
definite matrix $M \in \mathbb{R}^{k \times k}$ with $\ttr M = m \in
\mathbb{N}$ as $M = AA^t$, where $A$ has unit norm columns.

%% Finally, in Lemma \ref{lemma:lax} we generalize a lemma for
%% calculating $\frac{\der}{\der t} \log \det (I + X(t))$, where
%% $X(t)$ is an operator valued function \cite{Lax07}.

### 2023-05-14 latex/paper.tex
Done?

\author{Yair Daon\thanks{Porter School of the Environment and Earth
    Sciences, Tel Aviv University, Tel Aviv, Israel
    (\email{yair.daon@gmail.com}).}}
%% \address{Porter School of the Environment and Earth
%%   Sciences, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{Porter School of Environment Studies, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{}
%% \email{yair.daon@gmail.com}
% The 2010 edition of the Mathematics Subject Classification is the
% current definitive version.
%% \subjclass{Primary: 
%%   62F15, % Statistics - bayesian inference
%%   35R30, % PDE - Inverse problems
%%   Secondary:
%%   28C20,
%%   % Measure and integration - set functions and measures and integrals
%%          % in infinite-dimensional spaces (Wiener measure, Gaussian
%%   % measure, etc.
%% }

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  Inversion for parameteres of physical processes typically requires
  taking expensive measurements. Finding an optimal set of
  measurements is known as the optimal design problem. Surprisingly,
  measurement locations in optimal design are sometimes extremely
  clustered. Practically and intuitively, measurement clusterization
  is a huge problem. Researchers usually avoid measurement
  clusterization by somehow modifying the optimal design problem. We
  consider a certain flavor of the optimal design problem, based on
  the Bayesian D-optimality criterion. We suggest an analytically
  tractable model for Bayesian D-optimal designs in linear inverse
  problems over Hilbert spaces. We prove that measurement
  clusterization poses no obstruction to D-optimality. We demonstrate
  that measurement clusterization is a generic property of D-optimal
  designs, and prove that measurement correlation mitigates the
  clusterization of measurements. We also give a full characterization
  of D-optimal designs, which shows that D-optimal designs uniformly
  reduce uncertainty in the space of observations. Most importantly,
  we show how measurement clusterization can be avoided without
  changing the optimal design problem.
\end{abstract}

@@

\section{Acknowledgements}
This study is a part of my PhD thesis. Most of it was written under
the instruction of Prof. Georg Stadler of the Courant Institute of
Mathematical Sciences. I would like to thank Prof. Stadler for his
great mentorship, attention to details and kind behavior.

This research was supported in part by an appointment with the
National Science Foundation (NSF) Mathematical Sciences Graduate
Internship (MSGI) Program sponsored by the NSF Division of
Mathematical Sciences. This program is administered by the Oak Ridge
Institute for Science and Education (ORISE) through an interagency
agreement between the U.S. Department of Energy (DOE) and NSF. ORISE
is managed for DOE by ORAU. All opinions expressed in this paper are
the author's and do not necessarily reflect the policies and views of
NSF, ORAU/ORISE, or DOE.  This work was also supported by The Raymond
and Beverly Sackler Post-Doctoral Scholarship.

\input{appendix}

>>>>

%% \author{Yair Daon\thanks{Porter School of the Environment and Earth
%%     Sciences, Tel Aviv University, Tel Aviv, Israel
%%     (\email{yair.daon@gmail.com}).}}
\author{Yair Daon\thanks{Azrieli Faculty of Medicine, Bar Ilan
    University, Safed, Israel.  (\email{yair.daon@gmail.com}).}}

%% \address{Porter School of the Environment and Earth
%%   Sciences, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{Porter School of Environment Studies, Tel Aviv University\\ Tel Aviv, Israel}
%% %% \curraddr{}
%% \email{yair.daon@gmail.com}
% The 2010 edition of the Mathematics Subject Classification is the
% current definitive version.
%% \subjclass{Primary: 
%%   62F15, % Statistics - bayesian inference
%%   35R30, % PDE - Inverse problems
%%   Secondary:
%%   28C20,
%%   % Measure and integration - set functions and measures and integrals
%%          % in infinite-dimensional spaces (Wiener measure, Gaussian
%%   % measure, etc.
%% }

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  Inversion for parameteres of physical processes typically requires
  taking expensive measurements; finding an optimal set of
  measurements is known as the optimal design problem. Surprisingly,
  measurement locations in optimal designs are sometimes extremely
  clustered. Since measurement clusterization is nonintuitive,
  researchers often avoid it by modifying the optimal design
  problem. We consider a certain flavor of the optimal design problem,
  based on the Bayesian D-optimality criterion, and suggest an
  analytically tractable model for Bayesian D-optimal designs in
  linear inverse problems over Hilbert spaces. We prove that
  measurement clusterization poses no obstruction to D-optimality. We
  demonstrate that measurement clusterization is a generic property of
  D-optimal designs, and prove that correlated noise between
  measurements mitigates clusterization. We also give a full
  characterization of D-optimal designs, which shows that D-optimal
  designs uniformly reduce uncertainty in the space of
  observations. Finally, we show how measurement clusterization can be
  avoided without changing the optimal design problem.
\end{abstract}

@@

\section{Acknowledgements}
This study is a result of research I started during my PhD studies
under the instruction of Prof. Georg Stadler of the Courant Institute
of Mathematical Sciences. I would like to thank him for his great
mentorship, attention to details and kind behavior. This work was
supported by The Raymond and Beverly Sackler Post-Doctoral
Scholarship.

This research was supported in part by an appointment with the
National Science Foundation (NSF) Mathematical Sciences Graduate
Internship (MSGI) Program sponsored by the NSF Division of
Mathematical Sciences. This program is administered by the Oak Ridge
Institute for Science and Education (ORISE) through an interagency
agreement between the U.S. Department of Energy (DOE) and NSF. ORISE
is managed for DOE by ORAU. All opinions expressed in this paper are
the author's and do not necessarily reflect the policies and views of
NSF, ORAU/ORISE, or DOE.

%% \input{appendix}
